\texttt{aeons}: approximating the end of nested sampling

### 2024-07-17 review_response/paper.tex
second minor revision for journal submission

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the case of the $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
\begin{equation}\label{eq:logL_x}
    \log\Like = \logLmax - \frac{|\bm{x}|^{2}}{2\sigma^{2}}, \quad \bm{x} \in \mathbb{R}^{d}.
\end{equation}
This case is common and useful because many likelihoods are near-Gaussian, and most nested sampling implementations require the prior to be specified in terms of a transformation into the unit hypercube i.e. in prior probability units. Providing that $\sigma^d \ll 1$, the likelihood and posterior can be approximated as
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2,
\end{equation}
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior volumes are uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
\end{equation}

@@

We can also put compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have

@@

\subsubsection*{Comparisons}
\cref{fig:beta_posteriors} shows the posterior distribution as a function of $\log X$ for a 32-d spherical Gaussian when set to the above four temperatures, as known at a given iteration roughly 30\% through a run. Lower prior volumes are to the right, and 25 samples are taken from the prior volume distribution to capture the nested sampling uncertainties. When $\beta = 1$, the untransformed posterior is concentrated at the single point with the highest likelihood. This is no longer the case when $\beta$ is set to any of the other values, causing the posterior to have non-vanishing variance. The termination temperature can be seen to transform the posterior to have exactly $\epsilon$ of the total evidence to the right hand side of the current iteration (dotted black line). $\beta_\mathrm{M}$ makes the peak of the posterior with respect to $\log X$ coincide with the current iteration, and $\beta_\mathrm{C}$ produces a very similar result from its own criteria. 25 samples are taken from the distribution of the Bayesian temperature and the resultant posterior plotted for each, displaying a much broader range of uncertainty compared to the other plots.
\par
\cref{fig:beta_plots} then shows how these four temperatures evolve during the run (i.e. as a function of $\log X$) for two cases: one containing a classic slab-spike phase transition and one without. The behaviour of the former case is especially useful, because it is well-known to trouble temperature-based samplers. 
\par
In the case of a single phase all temperatures are consistent, with the exception that this only holds for $\epsilon \approx 0.5$ for the termination temperature. This is expected, since we can see from \cref{fig:beta_posteriors} that this choice of epsilon places the current iteration roughly at the peak of the posterior just as $\beta_M$ does. Other choices of $\epsilon$ shift the profile up and down. This user-dependence can offer more flexibility, but also be a disadvantage as an extra parameter to be tuned. The temperatures diverge, however, during a phase transition. Notably, with the exception of the canonical temperature they all \textit{decrease} then increase again, in contrast to a typically monotonic annealing schedule. It is also evident from the plots that $\beta_\mathrm{B}$ is the only temperature that incorporates a measure of uncertainty.
\begin{figure*}
\begin{center}
    \includegraphics{figures/beta_posteriors.pdf}
\end{center}
\caption{The posterior distribution at an intermediate point 30\% of the way through a nested sampling run for a 32$d$ spherical Gaussian, with the likelihood transformed from  $\Like \to \Like^{\beta}$ for several different choices of $\beta$. The untransformed posterior has all its weight at a single point. The other $\beta$ produce relatively similar transformations with finite posterior variance. Notably, the termination temperature is dependent on a manual choice of $\epsilon$, and the Bayesian temperature includes uncertainty in $\beta$ itself.}
\label{fig:beta_posteriors}
\end{figure*}

\begin{figure}
\begin{center}
    \includegraphics{figures/beta_plots.pdf}
\end{center}
\caption{Inferred temperatures using the termination ($\epsilon = 10^{-3}$), microcanonical, canonical and Bayesian definitions. The shaded region shows the $1-2\sigma$ uncertainties for the Bayesian temperature. $\beta_\epsilon$ is offset from the others, but can be shifted up and down depending on the chosen $\epsilon$. The other $\beta$ are all are consistent in a single phase, but differ during a phase transition.}
\label{fig:beta_plots}
\end{figure}
Despite the above theoretical justifications, one should remember the original motivation for introducing $\beta$ was to utilise the extra degree of freedom it provides, to transform the likelihood without losing information. As we will see in \cref{sec:choosing_the_best_temperature}, one can choose the exact definition depending on what is useful for the problem at hand.

\subsection{Dimensionality}\label{sec:dimensionality}
We can now use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible. \citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality $\dG$, defined as the posterior variance of the information content:
\begin{equation}\label{eq:d_G}
    \frac{\dG}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
    = \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
\end{equation}
As is discussed in \citet{Handley_2019}, $\dG$ holds valuable information about the posterior, measuring the number of spatial degrees of freedom in the posterior in the manner of principal component analysis, but without introducing arbitrary cut-off eigenvalues. It is exact in the case of a multivariate Gaussian likelihood that is small in length scale relative to the prior, so for non-Gaussian likelihoods $\dG$ can be considered the equivalent dimension if it were actually a Gaussian. Furthermore, it can be written as the posterior variance of the log-likelihood:
\begin{equation}
    \frac{\dG}{2} = \mathrm{Var}(\log\Like)_\mathcal{P}.
\end{equation}
Unfortunately, calculating this quantity in the middle of a run tends to return zero. That naturally follows from the results of \cref{sec:logL}, which shows that an intermediate set of samples typically have all their weight concentrated at a single point, the maximum live point, and a single point has vanishing variance. This behaviour continues practically until the run is close to termination. Re-weighting the likelihood by re-setting the temperature, however, leads this to no longer be the case. As shown in \cref{fig:beta_posteriors}, the posterior variance (and hence dimensionality) is zero at $\beta = 1$ but nonzero for the various other temperatures presented. 
\par
We find that the $\dG$ calculated from these transformed posteriors are exactly the Bayesian model dimensionalities of the \textit{true} posterior. \cref{fig:d_G_gaussians} demonstrates this using the microcanonical and Bayesian temperatures, on several spherical Gaussians for which the true value of $\dG$ is known; except for the earliest iterations, the right value is found to within standard error for all the likelhoods. The Bayesian temperature has slightly larger error bounds, as a more cautious estimate, but otherwise the estimates are consistent. 
\par
This provides a crucial piece of information, allowing us to access the Gaussian dimensionality of the true posterior before the run is complete. Before we discuss how this is can be used to aid endpoint estimation, we consider how the prior can affect the value of $\dG$ mid-run.  
\begin{figure*}
\begin{center}
    \includegraphics{figures/d_G_gaussians.pdf}
\end{center}
\caption{Bayesian model dimensionalities estimated throughout a run for spherical Gaussians of varying dimensions, which were not previously accessible before the end of a run because the posterior samples have vanishing variance. Calculating the value from the transformed likelihood $\Like \to \Like^{\beta}$, however, recovers the dimensionality of the final posterior before it is reached, which is shown in the above plots. Shading represents the $1-2\sigma$ uncertainties on the estimates $\hat{d}_\mathrm{G}$ of the true posterior dimensionality. Only the microcanonical and Bayesian temperatures are shown for clarity of comparison.}
\label{fig:d_G_gaussians}
\end{figure*}

@@

The key point is that in panel B, the live point cloud is actually one-dimensional. When comapred to a truly one-dimensional posterior (\cref{fig:d_G_elongated_1D}), at point B the samples look the same in all aspects. The consequence is that at point B, we fundamentally cannot know if there are more parameters `waiting' to be constrained, or if we have already found all the relevant directions. There are no clever inference tricks to get around it, in the same way that we cannot infer a `spike' from a slab and spike posterior when all the samples have yet to find the `spike'. Knowing the total number of parameters of the prior and likelihood to get the `parameter dimensionality' also isn't helpful, because the posterior dimensionality could be significantly different due to parameter degeneracies or non-Gaussian likelihoods.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, which depends on the live point schedule chosen by the user. The second term is the average time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation. The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior. This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principal interest in this section.
\par
Our aim is therefore first to quantify the amount of compression required to reach the endpoint at approximately $\log X = -\DKL$, which provides an estimate of the how far the sampler still needs to penetrate into the posterior. Once the required compression $\log X_\mathrm{f}$ is known, multiplying by the harmonic mean of the number of live points through the run gives the iteration count. Converting the number of nested sampling iterations to a run-time requires knowledge of how the sampling efficiency changes throughout the run, which we will not attempt to do because the variation is highly implementation-specific. 
\par
On the final point: the iteration progress is roughly proportional to the runtime progress when the sampling efficiency does not vary in order of magnitude. When that is not the case, the estimate will still be useful if the variation is known \textit{a priori}. For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.

@@

Previous work, such as the fitting procedure in the original Nelder-Mead simplex algorithm to estimate the Hessian \citep{nelder_mead}, and the BAMBI nested sampling paper \citep{Graff_2012}, suggest fast Gaussian approximations can be a good candidate for (online) modelling of likelihoods. The following method uses a similar approach, but unlike the previous schemes do not fit directly in parameter space.

@@

\subsection{Choosing the best temperature}\label{sec:choosing_the_best_temperature}
We estimate the endpoint using the above procedure for three previously discussed test likelihoods, each in a unit hypercube prior:
\begin{enumerate}[leftmargin=*]
    \item Spherical Gaussian; $\sigma = 10^{-3}$
        \begin{equation}
        \log \Like = -\frac{|\bm{x}|^2}{2\sigma^2}, \quad \bm{x} \in \mathbb{R}^{32}
        \end{equation}
    \item Elongated Gaussian, two well-separated length scales 
        \begin{equation}
            \log\Like = -\frac{1}{2} \bm{x}^\intercal \Sigma^{-1}\bm{x}, \quad \bm{x} \in \mathbb{R}^6
        \end{equation}
        \begin{equation*}
            \Sigma_\mathrm{separated} = \mathrm{diag}(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}),
        \end{equation*}
    \item Elongated Gaussian, gradually changing length scales
        \begin{equation}
            \Sigma_\mathrm{close} = \mathrm{diag}(10^{-2}, 10^{-3}, 10^{-4}, 10^{-5}, 10^{-6}, 10^{-7}).
        \end{equation}
\end{enumerate}
Results are shown in \cref{fig:beta_endpoint_comparison}. Broadly speaking, the performance is similar, which is expected since the temperatures themselves are relatively consistent with each other (\cref{fig:beta_plots}).
\par
No estimates can be calculated for the termination temperature very early on in the run, because there is no choice of $\beta$ that will make the current iteration the termination point. For example, this is certainly true when $X_i > \epsilon$.
\par
We choose to use the Bayesian temperature in our implementation from this point onwards, because it incorporates the extra degree of uncertainty in the temperature itself, and this broader account of uncertainty is helpful when we are estimating endpoints using limited information. However, the choice is to some degree one of personal taste, and it remains reasonable to make estimates with the others.
\begin{figure*}
\subfloat{\includegraphics{figures/beta_endpoint_comparison_gauss_32.pdf}\label{gauss_32}}
\newline
\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d.pdf}\label{elongated_gaussian_separated}}
\newline
\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d_close.pdf}\label{elongated_gaussian_close}}
\caption{Endpoint estimates obtained from the procedure outlined in \cref{sec:how_to_extrapolate}, using different $\beta$ to regularize the $\dG$. The true endpoints are shown in dotted red, while the estimates with their $1-2\sigma$ uncertainties are shaded in blue. Progress is measured in terms of the compression $-\log X$ along the x-axis, while the estimates on the y-axis are in terms of the predicted final compression $-\log \hat{X}_\mathrm{f}$. The estimates are broadly consistent, but notable differences include the failure of the termination temperature to obtain estimates at early iterations, and the larger uncertainty bounds of the Bayesian temperature.}
\label{fig:beta_endpoint_comparison}
\end{figure*}

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity.
\subsubsection{Integral progress}
An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
This has a several shortcomings. First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run. Finally, approximation \cref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum. This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.

>>>>

\subsection{Log-likelihood}\label{sec:logL}
We now consider the distribution of the samples in log-likelihood. To get an insight into the analytics, we will examine the case of the isotropic, $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
\begin{equation}\label{eq:logL_x}
    \log\Like = \logLmax - \frac{|\bm{x}|^{2}}{2\sigma^{2}}, \quad \bm{x} \in \mathbb{R}^{d}.
\end{equation}
This case is common and useful because many likelihoods are near-Gaussian, and most nested sampling implementations require the prior to be specified in terms of a transformation into the unit hypercube i.e. in prior probability units. Providing that $\sigma^d \ll 1$, the likelihood and posterior can be approximated as
\begin{equation}\label{eq:gaussian_logL}
    \log \Like = \logLmax - X^{2/d}/2\sigma^2,
\end{equation}
\begin{equation}
    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
    \label{eqn:PX}
\end{equation}
which in terms of $\log\mathcal{L}$ becomes
\begin{equation}
    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the prior volumes are uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
\begin{equation}
\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
\end{equation}

@@

We can also compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have

@@

\subsubsection*{Comparisons}
\cref{fig:beta_posteriors} shows the posterior distribution as a function of $\log X$ for a 32-d isotropic Gaussian when set to the above four temperatures, as known at a given iteration roughly 30\% through a run. Lower prior volumes are to the right, and 25 samples are taken from the prior volume distribution to capture the nested sampling uncertainties. When $\beta = 1$, the untransformed posterior is concentrated at the single point with the highest likelihood. This is no longer the case when $\beta$ is set to any of the other values, causing the posterior to have non-vanishing variance. The termination temperature can be seen to transform the posterior to have exactly $\epsilon$ of the total evidence to the right hand side of the current iteration (dotted black line). $\beta_\mathrm{M}$ makes the peak of the posterior with respect to $\log X$ coincide with the current iteration, and $\beta_\mathrm{C}$ produces a very similar result from its own criteria. 25 samples are taken from the distribution of the Bayesian temperature and the resultant posterior plotted for each, displaying a much broader range of uncertainty compared to the other plots.
\par
\cref{fig:beta_plots} then shows how these four temperatures evolve during the run (i.e. as a function of $\log X$) for two cases: one containing a classic slab-spike phase transition and one without. The behaviour of the former case is especially useful, because it is well-known to trouble temperature-based samplers. 
\par
In the case of a single phase all temperatures are consistent, with the exception that this only holds for $\epsilon \approx 0.5$ for the termination temperature. This is expected, since we can see from \cref{fig:beta_posteriors} that this choice of epsilon places the current iteration roughly at the peak of the posterior just as $\beta_M$ does. Other choices of $\epsilon$ shift the profile up and down. This user-dependence can offer more flexibility, but also be a disadvantage as an extra parameter to be tuned. The temperatures diverge, however, during a phase transition. Notably, with the exception of the canonical temperature they all \textit{decrease} then increase again, in contrast to a typically monotonic annealing schedule. It is also evident from the plots that $\beta_\mathrm{B}$ is the only temperature that incorporates a measure of uncertainty.
\begin{figure*}
\begin{center}
    \includegraphics{figures/beta_posteriors.pdf}
\end{center}
\caption{The posterior distribution at an intermediate point 30\% of the way through a nested sampling run for a 32$d$ isotropic Gaussian, with the likelihood transformed from  $\Like \to \Like^{\beta}$ for several different choices of $\beta$. The untransformed posterior has all its weight at a single point. The other $\beta$ produce relatively similar transformations with finite posterior variance. Notably, the termination temperature is dependent on a manual choice of $\epsilon$, and the Bayesian temperature includes uncertainty in $\beta$ itself.}
\label{fig:beta_posteriors}
\end{figure*}

\begin{figure}
\begin{center}
    \includegraphics{figures/beta_plots.pdf}
\end{center}
\caption{Inferred temperatures using the termination ($\epsilon = 10^{-3}$), microcanonical, canonical and Bayesian definitions. The shaded region shows the $1-2\sigma$ uncertainties for the Bayesian temperature. $\beta_\epsilon$ is offset from the others, but can be shifted up and down depending on the chosen $\epsilon$. The other $\beta$ are all consistent in a single phase, but differ during a phase transition.}
\label{fig:beta_plots}
\end{figure}
Despite the above theoretical justifications, one should remember the original motivation for introducing $\beta$ was to utilise the extra degree of freedom it provides, to transform the likelihood without losing information. As we will see in \cref{sec:choosing_the_best_temperature}, one can choose the exact definition depending on what is useful for the problem at hand.

\subsection{Dimensionality}\label{sec:dimensionality}
We can now use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible. \citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality $\dG$, defined as the posterior variance of the information content:
\begin{equation}\label{eq:d_G}
    \frac{\dG}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
    = \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
\end{equation}
As is discussed in \citet{Handley_2019}, $\dG$ holds valuable information about the posterior, measuring the number of spatial degrees of freedom in the posterior in the manner of principal component analysis, but without introducing arbitrary cut-off eigenvalues. It is exact in the case of a multivariate Gaussian likelihood that is small in length scale relative to the prior, so for non-Gaussian likelihoods $\dG$ can be considered the equivalent dimension if it were actually a Gaussian. Furthermore, it can be written as the posterior variance of the log-likelihood:
\begin{equation}
    \frac{\dG}{2} = \mathrm{Var}(\log\Like)_\mathcal{P}.
\end{equation}
Unfortunately, calculating this quantity in the middle of a run tends to return zero. That naturally follows from the results of \cref{sec:logL}, which shows that an intermediate set of samples typically have all their weight concentrated at a single point, the maximum live point, and a single point has vanishing variance. This behaviour continues practically until the run is close to termination. Re-weighting the likelihood by re-setting the temperature, however, leads this to no longer be the case. As shown in \cref{fig:beta_posteriors}, the posterior variance (and hence dimensionality) is zero at $\beta = 1$ but nonzero for the various other temperatures presented. 
\par
We find that the $\dG$ calculated from these transformed posteriors are exactly the Bayesian model dimensionalities of the \textit{true} posterior. \cref{fig:d_G_gaussians} demonstrates this using the microcanonical and Bayesian temperatures, on several isotropic Gaussians for which the true value of $\dG$ is known; except for the earliest iterations, the right value is found to within standard error for all the likelihoods. The Bayesian temperature has slightly larger error bounds, as a more cautious estimate, but otherwise the estimates are consistent. 
\par
This provides a crucial piece of information, allowing us to access the Gaussian dimensionality of the true posterior before the run is complete. Before we discuss how this can be used to aid endpoint estimation, we consider how the prior can affect the value of $\dG$ mid-run.  
\begin{figure*}
\begin{center}
    \includegraphics{figures/d_G_gaussians.pdf}
\end{center}
\caption{Bayesian model dimensionalities estimated throughout a run for isotropic Gaussians of varying dimensions, which were not previously accessible before the end of a run because the posterior samples have vanishing variance. Calculating the value from the transformed likelihood $\Like \to \Like^{\beta}$, however, recovers the dimensionality of the final posterior before it is reached, which is shown in the above plots. Shading represents the $1-2\sigma$ uncertainties on the estimates $\hat{d}_\mathrm{G}$ of the true posterior dimensionality. Only the microcanonical and Bayesian temperatures are shown for clarity of comparison.}
\label{fig:d_G_gaussians}
\end{figure*}

@@

The key point is that in panel B, the live point cloud really only varies in one dimension, despite the true posterior having two constrained parameters. When compared to another $2d$ posterior which truly only has one constrained parameter, with the other being flat (\cref{fig:d_G_elongated_1D}), we find that at point B the samples look the same in all aspects. The consequence is that at point B, we fundamentally cannot know if there are more parameters `waiting' to be constrained, or if we have already found all the relevant directions. There are no clever inference tricks to get around it, in the same way that we cannot infer a `spike' from a slab and spike posterior when all the samples have yet to find the `spike'. Knowing the total number of parameters of the prior and likelihood to get the `parameter dimensionality' also isn't helpful, because the posterior dimensionality could be significantly different due to parameter degeneracies or non-Gaussian likelihoods.

@@

\section{Endpoint prediction}\label{sec:endpoint}
As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
\begin{equation}
    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
\end{equation}
The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, which depends on the live point schedule chosen by the user. The second term is the average time per likelihood evaluation. The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation. The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior. This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principal interest in this section.
\par
Our aim is therefore first to quantify the amount of compression required to reach the endpoint at approximately $\log X = -\DKL$, which provides an estimate of how far the sampler still needs to penetrate into the posterior. Once the required compression $\log X_\mathrm{f}$ is known, multiplying by the harmonic mean of the number of live points through the run gives the iteration count. Converting the number of nested sampling iterations to a run-time requires knowledge of how the sampling efficiency changes throughout the run, which we will not attempt to do because the variation is highly implementation-specific. 
\par
On the final point: the iteration progress is roughly proportional to the runtime progress when the sampling efficiency does not vary in order of magnitude. When that is not the case, the estimate will still be useful if the variation is known \textit{a priori}. For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.

@@

Previous work, such as the fitting procedure in the original Nelder-Mead simplex algorithm to estimate the Hessian \citep{nelder_mead}, and the BAMBI nested sampling paper \citep{Graff_2012}, suggest fast Gaussian approximations can be a good candidate for (online) modeling of likelihoods. The following method uses a similar approach, but unlike the previous schemes does not fit directly in parameter space.

@@

\subsection{Choosing the best temperature}\label{sec:choosing_the_best_temperature}
We estimate the endpoint using the above procedure for three previously discussed test likelihoods, each in a unit hypercube prior:
\begin{enumerate}[leftmargin=*]
    \item Isotropic Gaussian; $\sigma = 10^{-3}$
        \begin{equation}
        \log \Like = -\frac{|\bm{x}|^2}{2\sigma^2}, \quad \bm{x} \in \mathbb{R}^{32}
        \end{equation}
    \item Elongated Gaussian, two well-separated length scales 
        \begin{equation}
            \log\Like = -\frac{1}{2} \bm{x}^\intercal \Sigma^{-1}\bm{x}, \quad \bm{x} \in \mathbb{R}^6
        \end{equation}
        \begin{equation*}
            \Sigma_\mathrm{separated} = \mathrm{diag}(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}),
        \end{equation*}
    \item Elongated Gaussian, gradually changing length scales
        \begin{equation}
            \Sigma_\mathrm{close} = \mathrm{diag}(10^{-2}, 10^{-3}, 10^{-4}, 10^{-5}, 10^{-6}, 10^{-7}).
        \end{equation}
\end{enumerate}
Results are shown in \cref{fig:beta_endpoint_comparison}. Broadly speaking, the performance is similar, which is expected since the temperatures themselves are relatively consistent with each other (\cref{fig:beta_plots}).
\par
No estimates can be calculated for the termination temperature very early on in the run, because there is no choice of $\beta$ that will make the current iteration the termination point. For example, this is certainly true when $X_i > \epsilon$.
\par
We choose to use the Bayesian temperature in our implementation from this point onwards, because it incorporates the extra degree of uncertainty in the temperature itself, and this broader account of uncertainty is helpful when we are estimating endpoints using limited information. However, the choice is to some degree one of personal taste, and it remains reasonable to make estimates with the others.
\begin{figure*}
\subfloat{\includegraphics{figures/beta_endpoint_comparison_gauss_32.pdf}\label{gauss_32}}
\newline
\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d.pdf}\label{elongated_gaussian_separated}}
\newline
\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d_close.pdf}\label{elongated_gaussian_close}}
\caption{Endpoint estimates obtained from the procedure outlined in \cref{sec:how_to_extrapolate}, using different $\beta$ to regularize the $\dG$. The true endpoints are shown in dotted red, while the estimates with their $1-2\sigma$ uncertainties are shaded in blue. Progress is measured in terms of the compression $-\log X$ along the x-axis, while the estimates on the y-axis are in terms of the predicted final compression $-\log \hat{X}_\mathrm{f}$. The estimates are broadly consistent, but notable differences include the failure of the termination temperature to obtain estimates at early iterations, and the larger uncertainty bounds of the Bayesian temperature.}
\label{fig:beta_endpoint_comparison}
\end{figure*}

\subsection{Alternative approaches}
More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented. However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted. Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity.
\subsubsection{Integral progress}
An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
\begin{equation}\label{eq:integral_progress}
    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
\end{equation}
This has several shortcomings. First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width. Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run. Finally, approximation \cref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum. This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.

\subsubsection{Extrapolating evidence increments}
Seasoned watchers of nested sampling runs might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges. At an intermediate stage of the run, the most recent outputs look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate} (which in this case was a real run for a isotropic Gaussian with $d = 16$, $\sigma = 0.01$). The right hand side of \cref{fig:inc_extrapolate} plots the gain in log-evidence at every 500 iterations, then extrapolates them with a linear and exponential profile. This is then used to calculate the point at which the gain is so small that the termination condition is met, shown separately for the linear and exponential extrapolations.
\par
Compared to the true endpoint (found by completing the run), the linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile. The increments are also not exactly exponential, since the exponential fit leads to a large over-prediction. The predicted endpoint over the course of a run, as shown in \cref{fig:inc_predictions}, shows the same result. One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
\par
More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease. Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
\begin{figure}
\begin{adjustbox}{valign=t, scale=0.7}
\subfloat{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $\Delta\log Z$   \\
\hline
5000 & -1435.8 & 190.8 \\
5500 & -1264.6 & 171.2 \\
6000 & -1123.7 & 140.9 \\
6500 & -991.5 & 132.2 \\
7000 & -885.0 & 106.6 \\
7500 & -790.3 & 94.7 \\
8000 & -702.6 & 87.7 \\
8500 & -619.7 & 82.9 \\
9000 & -551.8 & 67.9 \\
9500 & -492.7 & 59.1 \\
\hline
\end{tabular}}
\end{adjustbox}
\quad
\begin{adjustbox}{valign=t, scale=0.9}
    \subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
\end{adjustbox}
\caption{Extrapolating the increases in evidence $\Delta \log \mathcal{Z}$ to predict the endpoint. The left column shows the most recent ten outputs from a nested sampling run currently at $n_\mathrm{dead} = 10,000$, together with the difference every 500 iterations, which appear to converge at a steady rate. The right hand side plots these past 10 increments (blue) as a function of the compression $-\log X_\mathrm{f}$, and attempts to extrapolate them via a linear and exponential profile (pink). This is used to predict when the increment will be small enough for the termination condition to be met (dashed pink). The true progress of the increments, obtained after completing the run, is plotted in black.}
\label{fig:inc_extrapolate}
\end{figure}
\begin{figure}
\includegraphics{figures/inc_predictions.pdf}
\caption{Endpoint predictions for a isotropic Gaussian, with $1-2\sigma$ uncertainties shaded. Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which perform considerably worse than the method of extrapolating the likelihood.}
\label{fig:inc_predictions}
\end{figure}
